MS Technical Paper: \\ Placement Algorithms for Heterogeneous FPGAs

### 2025-07-21 latex/paper/sections/11-placement_results.tex
cooling schedules

{
    \centering
    \includegraphics[width=0.6\columnwidth]{figures/placement/cooling_schedule_10000_98.png}
    \captionof{figure}{Cooling Schedule: \(T_0=10000\), \(\alpha=0.98\)}
    \label{fig:cooling_schedule_10000_98}
}

\newpage
Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.

@@

The tangle of red lines represent the nets between the \texttt{Site}s, which are yet to be routed at this stage. 
Brighter red lines represent nets of higher total HPWL, while duller red lines represent nets of lower total HPWL.

Notice there are two brightly red nets in every figure that have a particularly high fan-out:
one sourced from the near-center, and another sourced from the top right of the device.

The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock trees. 
Since clock nets tend to have very high fanout, they can use up a significant portion of the routing resources if they are routed through general routing.
Modern FPGAs like the 7-Series have dedicated clock routes which can efficiently distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
These special clock routes are accessed by routing the raw clock signal from the crystal oscillator IO port to one these on-chip clock buffers which connects to the on-chip clock tree network.
It is always best practice to use them to improve the routability of the design, and simultaneously, to reduce clock skew and improve timing. 
Our simple FIR Filter design only operates on a single clock domain, so only one clock buffer \texttt{Site} is used. 
An HDL design operating on many clock domains will use more clock buffers or a dedicated clock generator such as a phase locked loop (PLL) or mixed-mode clock manager (MCMM).

@@

Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the large grid of \texttt{SLICEL}s (cyan).

@@

>>>>

{
    \centering
    \includegraphics[width=0.6\columnwidth]{figures/placement/cooling_schedule_10000_98.png}
    \captionof{figure}{Cooling Schedule: \(T_0=10000\), \(\alpha=0.98\)}
    \label{fig:cooling_schedule_10000_98}
}
\vspace{1.0cm}

Figures \ref{fig:PGRSnapshots}, \ref{fig:PGMSnapshots}, \ref{fig:PARSnapshots}, \ref{fig:PAMSnapshots}, and \ref{fig:PAHSnapshots} show snapshots of the placement progress on the physical \texttt{device} after 0, 10, 100, and 300 passes for each placer variant.

@@

The tangle of red lines represent the nets between the \texttt{Site}s, which are yet to be routed at this stage. 
Brighter red lines represent nets of higher total HPWL, while duller red lines represent nets of lower total HPWL.
Notice there are two brightly red nets in every figure that have a particularly high fanout:
one sourced from the near-center, and another sourced from the top right of the device.

The singular white box in the near-center of all of the figures represents an occupied \texttt{BUFGCTRL} \texttt{Site}, which is a clock buffer used to route clock nets. 
Since clock nets tend to have very high fanout, they can use up an unnecessarily large portion of routing resources if they are routed through general routing.
Modern FPGAs like the 7-Series have dedicated clock tree routes which can evenly distribute clock signals to all logical \texttt{Site}s without burdening the general routing resources.
These special clock routes are accessed by routing the raw clock signal from the crystal oscillator IO port to one these on-chip clock buffers which connects to the on-chip clock tree network.
It is always best practice to use them to improve the routability of the design, and simultaneously, to reduce clock skew and improve timing. 
Our simple FIR Filter design only operates on a single clock domain, so only one clock buffer \texttt{Site} is used. 
An HDL design operating on many clock domains will use more clock buffers or a dedicated clock generator such as a phase locked loop (PLL) or mixed-mode clock manager (MCMM).

@@

Also note the sparsity of the \texttt{DSP48E1} (yellow) and \texttt{RAMB18E1} (magenta) \texttt{Site} columns compared to the more uniform grid of \texttt{SLICEL}s (cyan), which poses a computational challenge for the legalization process in directed placement.

@@

\begin{multicols}{2}
    In addition to variations on the basic SA algorithm, we also test various cooling schedules on the basic SA algorithm.
    Shown in figure \ref{fig:VaryCoolingSchedules} are the cost histories for various cooling schedules, grouped three different initial temperatures \texttt{(5000, 10000, and 15000)}. 
    For each initial temperature group, we test four different cooling rates \texttt{(0.92, 0.94, 0.96, and 0.98)}, with the darkness of each curve corresponding with a slower cooling rate.
\end{multicols}

{
    \centering
    \includegraphics[valign=t, scale=0.55]{figures/results/cost_history_5000.png}
    \includegraphics[valign=t, scale=0.55]{figures/results/cost_history_10000.png}
    \includegraphics[valign=t, scale=0.55]{figures/results/cost_history_15000.png}
    \includegraphics[valign=t, scale=0.55]{figures/results/combined_cost_history_cooling.png}
    \captionof{figure}{\texttt{PlacerAnnealRandom} with various cooling schedules.}
    \label{fig:VaryCoolingSchedules}
}

### 2025-07-21 latex/paper/sections/12-future_work.tex
citations
cooling schedules

\section{Future Work}
In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} homogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area.

There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow.

The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.

>>>>

\section{Future Work}

\subsection{Expand Library of Primitives}
In this paper we have made many simplifications to the problem space just to make the placer easier to implement.
For example, our placer in its current state does not take advantage of \texttt{SLICEL}/\texttt{SLICEM} homogeneity and simply maps all SLICE \texttt{SiteInst}s onto \texttt{SLICEL}s. 
Recall that the SLICE Sites in Xilinx FPGAs typically come in a 75-25\% split between \texttt{SLICEL}s and \texttt{SLICEM}s. 
This means that we have rendered about 25\% of the CLB fabric unusable which will inevitably hurt wirelength minimization during placement since the \texttt{SiteInst}s must be spread over a larger area. 
Enabling \texttt{SLICEL}-\texttt{SLICEM} homogeneity can lead do greater logic density and consequently less total HPWL, but can make the packing process more complex and may contribute to higher routing congestion.

\subsection{Robustness}
In its current state, the prepacker and packer struggle with data busses larger than 24-bits, especially with DSP functions like multiplication. 
In such designs, the synthesizer will synthesize \texttt{CARRY4} chains with particular \texttt{EDIFHierPortInst} configurations that are currently not supported by our packer which eventually lead to errors or failures in the subsequent routing stage.
Further work is required to resolve these routing constraints.

\subsection{Improvements to Packing}
There is also no rule saying we must pack \texttt{Cells} into \texttt{SiteInst}s before placement or that we follow a strict prepacking-packing-placement flow.

Our current pack-place approach is a \texttt{Site}-centric approach and resembles that of Xilinx ISE, which is the predecessor to the Vivado design suite.
The latest editions of Vivado performs BEL-centric placement without necessarily locking \texttt{Cell}s into \texttt{Site}s, allowing a higher granularity of movement of \texttt{Cell}s during placement.

{
    \centering
    \includegraphics[width=\columnwidth]{figures/future_work/legalization.png}
    \captionof{figure}{Representative FPGA placement and packing flows. Figure taken from Wuxi et al. (2019), page 1 \cite{ExplicitPacking}}
}
\vspace{0.25cm}

\subsection{Improvements to SA}

\subsection{Force-Directed and Analytical Placement}
